---
title: "Gov 50 Final Project"
author: Adithya Madduri
description: "Analysis of impact of Teacher Pay on Student SAT scores for Massachusetts Public High Schools in 2017"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to many different political topics. I would be especially interested in examining data related to controversial United States Presidential elections from the past such as the election of 2000 or even more historical elections depending on how far back the data we have extends too. I'm also interested in looking at developing countries and possibly using Hans Rosling's Gapminder data set from the book factfullness. I think it would be cool to take a topic form the book and conduct some more statistical analysis on it and draw conclusions from that data. Overall, I'm interested in these two topics, but I'm open to a range of different ideas.


## Milestone number 2

Do higher teacher salaries contribute to enhanced student academic outcomes? In this study, I plan to examine how higher teacher salaries contribute to student academic outcomes. I hypothesize that higher teacher salaries will contribute to better student academic outcomes since teachers will be more committed to educating their students, contributing to better classroom instruction, and therefore better educational outcomes for the students. When teachers are paid less, I believe that they will be less committed to their jobs, and this will mean that students receive a lower quality of education and will likely mean that the future educational outcomes of the students are worse. The data source I’m using is the Massachusetts Public School Data for the year 2017, and my sample size is 1800 Public schools in Massachusetts, including elementary, middle, and high schools. My unit of analysis is an individual school, for example Abington high school. The explanatory variables of interest are average salary and total salary. The variable will be measured in United States Dollars. My outcome variables include various measures of academic performance, specifically average SAT Math and Reading scores for students from the school and the number of high school graduates from the school. For instance, if the average salary variable was equal to $45,000, that means that the average teacher salary was $45,000. Likewise for the outcome variable, if the average  SAT Math score for a school is 700, this corresponds to This data is taken from the Massachusetts Public School data for the year 2017 and its validity is established by the fact that it comes directly from the Massachusetts Department of Elementary and Secondary Education. If I observe higher average SAT scores and high school graduation rates when teacher pay is higher, then that will mean that higher teacher pay does correspond to better educational outcomes for students. However, if I observe that average SAT scores and graduation rates are not significantly higher when teacher pay is higher, then that will mean that teacher pay does not have a significant impact on educational outcomes. When I run a regression of average SAT scores and high school graduation rates on average teacher salary as an indicator variable a positive, significant coefficient would indicate support of my hypothesis.


## Milestone 3

```{r}
library(tidyverse)
library(readr)
data_final <- read_csv("MA_Public_Schools_2017.csv")
View(data_final)

```
```{r}
data_highschool <- data_final |>
  filter(`12_Enrollment` >= 1 | `11_Enrollment` >= 1 | `10_Enrollment` >= 1 | `9_Enrollment` >= 1)
```
```{r}
library(ggplot2)

data_highschool <- data_highschool |>
  mutate(Average_SAT = `Average SAT_Reading` + `Average SAT_Writing` + `Average SAT_Math`)
ggplot(data = data_highschool,
       mapping = aes(x = `Average In-District Expenditures per Pupil`,
                     y = Average_SAT)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)
```
```{r}
ggplot(data = data_highschool,
       mapping = aes(x = `Salary Totals`,
                     y = Average_SAT)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10()
```
```{r}
ggplot(data = data_highschool,
       mapping = aes(x = `Average Salary`,
                     y = Average_SAT)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10()
```


## Milestone 4
```{r}
library(tidyverse)
first_tibble <- data_highschool |>
  group_by(`Average Salary`) |>
  summarize(avg_SAT = mean(Average_SAT))

first_tibble



most_paid <- first_tibble |>
  slice_max(`Average Salary`, n = 5)

most_paid


least_paid <- first_tibble |>
  slice_min(`Average Salary`, n = 5)

least_paid
```
Looking at the data for the 5 highest and 5 lowest salary totals by district, we can see that SAT scores tend to be lower when salary totals are higher. From this data, for furtehr analysis, I need to clear out salary total values for districts that do not have an average SAT score. 

Looking at this tibble and the previous two charts, it appears that a higher salary total for teachers is correlated with lower SAT test scores. This can be seen from the negative line of best fit on the second scatter plot. The negative slope on the line means indicatest that the correlation between salary totals and SAT test scores is negative. Therefore, as teacher salary totals increase, we would expect to see lower SAT test scores. 

```{r}
fit <- lm(`Average Salary` ~ Average_SAT, data = data_highschool)
fit
coef(fit)
```
In the above code, I ran a linear regression so that we could get a clearer picture of the relationship between average Salary and Average SAT scores. As we can see from the coefficients for the linear regression, looking at all high schools in the state of Massachusetts for the year 2017, as average Salary increases the average SAT score for the district decreases by 4 points. This suggests a correlation, but does not necesarily imply a causual effect. There could be many other factors influencing student SAT scores outside of teacher pay. 

```{r}
data_highschool <- data_highschool |>
  mutate(
    pay_range = case_when(
      `Average Salary` < 65000 ~ "Under 65,000",
      `Average Salary` <=70000 & 'Average Salary' > 65000 ~ "Between 65,000 and 70,000",
      `Average Salary` <=75000 & 'Average Salary' > 70000 ~ "Between 70,000 and 75,000",
      `Average Salary` <=80000 & 'Average Salary' > 75000 ~ "Between 75,000 and 80,000",
      `Average Salary` <=85000 & 'Average Salary' > 80000 ~ "Between 80,000 and 85,000",
      `Average Salary` <=90000 & 'Average Salary' > 85000 ~ "Between 85,000 and 90,000",
      `Average Salary` <=95000 & 'Average Salary' > 90000 ~ "Between 90,000 and 95,000",
      `Average Salary` >=95000 ~ "Greater than 95,000"
    )
  )
data_highschool |>
  group_by(pay_range) |>  
  summarize(avg_SAT = mean(Average_SAT, na.rm = TRUE))

```
In this chart, we grouped the average salaries into 9 different categories. under $65,000; between $65,000 to 70,000; $70,000 to 75,000; $75,000 to 80,000; $80,000 to 85,000; $85,000 to 90,000; $90,000 to 95,000; and greater than $95,000. We then calculated the average SAT scores for each of these pay ranges. This again allows us to better understand the correlation between our two data columns. 




Results and Potential Explanations:

Possible Confounding Factors:

Conclusions:
From the data it is clear that there was a negative correlation between teacher salaries and average SAT scores for Public high schools in the state of Massachusetts for the year 2017. 
